---
title: "mrot diapause - lifespan"
author: "Karen M. Kapheim"
date: "`r Sys.Date()`"
output: pdf_document
---

This is analysis of the lifespan data for Megachile rotundata males & female 
that either did or did not diapause. These samples were provided as incubating 
cocoons from Lindsay McCabe (USDA PIRU).  

We changed out the cage style after we had some death in the original cage 
style. That is why there is a 'old_cage' variable. We want to see if the results
change when we include bees that spent time in the original cage design.

There were also some bees that had some discrepencies, missing data, or other 
issues, so made a column that excludes them. 

Thorax width was measured by Addie Lloyd in the Leica EZ Suite software. 

The original (locked) data sheet and pictures are in the folder 
'mrot_diapause_lifespan_immunity' on the KapheimLab google drive. 

The original hard copy of the data is in Kapheim's office.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
library(survival)
library(survminer)
library(ggfortify)
```

# Data

```{r data-1}
lifespan <- read_csv("MROT_lifespan_thorax_WorkingCopy.csv", 
                     col_select = c(1:10), 
                     col_names = TRUE,
                     col_types = cols(
                       Bee_ID = col_character(),
                       Diapause_NonDiapause = col_character(), 
                       Sex = col_character(), 
                       Date_emerged = col_date("%m/%d/%y"), 
                       Date_dead = col_date("%m/%d/%y"), 
                       Date_frozen = col_date("%m/%d/%y"), 
                       Thorax_length_mm = col_double(), 
                       Notes = col_character(), 
                       use_lifespan = col_character(), 
                       old_cage_design = col_character()
                     ))
problems(lifespan)
```

#### Make some new variables

We need a censor variable and a survival time.

```{r data-2}
lifespan <- lifespan %>% 
  filter(!Bee_ID == "NA") %>% 
  mutate(days_surv = as.numeric(Date_dead - Date_emerged), 
         status = as.logical("TRUE"), 
         combo = paste(Diapause_NonDiapause, Sex, sep = ""))
table(lifespan$combo)
lifespan$Diapause_NonDiapause = relevel(as.factor(lifespan$Diapause_NonDiapause), 
                                        ref = "ND")
```

#### Filter datasets

###### Only those without weird issues

```{r data-3}
lifespan.r <- lifespan %>% 
  filter(use_lifespan == "Y")
table(lifespan.r$combo)
```

###### Only those that spent whole life in new cage design

```{r data-4}
lifespan.r2 <- lifespan.r %>% 
  filter(old_cage_design == "N")
table(lifespan.r2$combo)
```

This eliminates most of the Diapause males. Maybe need to exclude males.

###### Only females

```{r data-5}
lifespan.r2F <- lifespan.r2 %>% 
  filter(Sex == "F")
table(lifespan.r2F$combo)
```

# Survival Analysis


#### Make a survival object

```{r surv-1}
km <- with(lifespan, Surv(days_surv, status))
km.r <- with(lifespan.r, Surv(days_surv, status))
km.r2 <- with(lifespan.r2, Surv(days_surv, status))
km.r2F <- with(lifespan.r2F, Surv(days_surv, status))
```

#### Get the stats

```{r surv-2}
km_fit <- survfit(Surv(days_surv, status) ~ combo, data=lifespan)
km_fit.r.combo <- survfit(Surv(days_surv, status) ~ combo, data=lifespan.r)
km_fit.r2.combo <- survfit(Surv(days_surv, status) ~ combo, data=lifespan.r2)
km_fit.r.DN <- survfit(Surv(days_surv, status) ~ Diapause_NonDiapause, 
                       data=lifespan.r)
km_fit.r2.DN <- survfit(Surv(days_surv, status) ~ Diapause_NonDiapause, 
                       data=lifespan.r2)
km_fit.r2F.DN <- survfit(Surv(days_surv, status) ~ Diapause_NonDiapause, 
                       data=lifespan.r2F)
km_fit.r.sex <- survfit(Surv(days_surv, status) ~ Sex, data=lifespan.r)

km_fit.r2.sex <- survfit(Surv(days_surv, status) ~ Sex, data=lifespan.r2)
```

#### Plots

###### All others

```{r surv-3}
# combo
autoplot(km_fit.r.combo)
ggsurvplot(km_fit.r.combo, data = lifespan.r, pval = TRUE)
# diapause
autoplot(km_fit.r.DN)
ggsurvplot(km_fit.r.DN, data = lifespan.r, pval = TRUE)
ggsurvplot(km_fit.r2.DN, data = lifespan.r2, pval = TRUE)
# sex
autoplot(km_fit.r.sex)
ggsurvplot(km_fit.r.sex, data = lifespan.r, pval = TRUE)
ggsurvplot(km_fit.r2.sex, data = lifespan.r2, pval = TRUE)
```

## Only females without issues

#### Summary data 

###### Min and max lifespan


```{r surv-4}
lifespan.r2F %>% 
  group_by(Diapause_NonDiapause) %>% 
  summarise(min = min(days_surv), 
            max = max(days_surv), 
            mean = mean(days_surv), 
            median = median(days_surv),
            sd = sd(days_surv),
            se = sd(days_surv)/sqrt(n()),
            n = n())

lifespan.r2F %>% 
  summarise(min = min(days_surv), 
            max = max(days_surv), 
            mean = mean(days_surv), 
            median = median(days_surv),
            sd = sd(days_surv),
            se = sd(days_surv)/sqrt(n()),
            n = n())
```

Min, Max, and Mean are similar, but median lifespan is 9 d longer (21%) for ND.

#### Body size difference between D and ND
```{r}
plot(density(lifespan.r2F$Thorax_length_mm, na.rm = TRUE))
shapiro.test(lifespan.r2F$Thorax_length_mm)

lifespan.r2F %>% 
  group_by(Diapause_NonDiapause) %>% 
  summarise(min = min(Thorax_length_mm), 
            max = max(Thorax_length_mm), 
            mean = mean(Thorax_length_mm), 
            median = median(Thorax_length_mm),
            sd = sd(Thorax_length_mm),
            se = sd(Thorax_length_mm)/sqrt(n()),
            n = n())

#Plot body size D vs ND
ggplot(lifespan.r2F, aes(x = Diapause_NonDiapause, y = Thorax_length_mm, fill = Diapause_NonDiapause, )) +
  geom_boxplot(alpha = 0.8, outlier.color=NA) +
  geom_jitter(position = position_jitter(width = 0.20)) +
  geom_signif(data = lifespan.r2F, stat = "signif", position = "identity", 
              comparisons = list(c("D", "ND")), map_signif_level = TRUE, 
              annotations = "***", textsize = 8) +
  coord_cartesian(ylim = c(1.8, 3.1)) +
  theme_classic() +
  scale_fill_manual(values = c("#D84727","#94D2BD")) +
  theme(legend.position = "none", 
        axis.title=element_text(size=16), axis.text=element_text(size=14)) +
  xlab("") + 
  scale_x_discrete(labels = c("Non-Diapause", "Diapause")) + 
  ylab("Intertegular Width (mm)")

ggsave("body_size_thorax_lifespan.png", width = 7, height = 5)

model <- lm(Thorax_length_mm ~ Diapause_NonDiapause, data = lifespan.r2F)
summary(model)
```
#### Summary of the survival curve

```{r surv-5}
print(km_fit.r2F.DN)
```

###### Plot

```{r surv-6}
autoplot(km_fit.r2F.DN)
ggsurvplot(km_fit.r2F.DN, data = lifespan.r2F, 
           pval = TRUE, 
           palette = c("#D84727", "#94D2BD"), 
           conf.int = TRUE, 
#           risk.table = "percentage", 
#           risk.table.col = "strata", 
           legend.title = "", 
           legend.labs = c("Non-diapause", "Diapause"), 
#           risk.table.height = 0.3, 
           break.time.by = 10, 
           xlab = "Time (days)", 
           surv.median.line = "hv", 
           font.x = c(14), 
           font.y = c(14)
           )
ggsave("survival.png", height = 4, width = 6)
```

# Cox-Hazard models

## All others

```{r cox-1}
# Fit a Cox proportional hazards model
fit.coxph.r <- coxph(km.r ~ Diapause_NonDiapause + Sex + Thorax_length_mm, 
                   data = as.data.frame(lifespan.r))
ggforest(fit.coxph.r, data = as.data.frame(lifespan.r))

```


```{r cox-2}
# Fit a Cox proportional hazards model
fit.coxph.r2 <- coxph(km.r2 ~ Diapause_NonDiapause + Sex + Thorax_length_mm, 
                   data = as.data.frame(lifespan.r2))
ggforest(fit.coxph.r2, data = as.data.frame(lifespan.r2))

```

## Females only

```{r cox-3}
# Fit a Cox proportional hazards model
fit.coxph.r2F <- coxph(km.r2F ~ Diapause_NonDiapause + Thorax_length_mm, 
                   data = as.data.frame(lifespan.r2F))
ggforest(fit.coxph.r2F, data = as.data.frame(lifespan.r2F), fontsize = 0.65)
ggsave("hazards.svg", height = 4.5, width = 6.5)
```

## book-keeping

```{r books}
sessionInfo()
citation("tidyverse")
citation("survival")
citation("survminer")
citation("ggfortify")
```
