---
title: "mrot diapause - oxidative stress"
author: "Karen M. Kapheim"
date: "`r Sys.Date()`"
output: pdf_document
---

This is an analysis of oxidative stress in newly emerged females who either 
went through diapause or developed directly. We used protein carbonylation in 
the thorax as a measure of oxidative stress.

## Set-up

```{r setup}
library(tidyverse)
library(nortest)
library(car)
library(lme4)
library(ggsignif)
```

## Get data

```{r data}
oxstress <- read_csv("MROT_protcarb_bodysize_data.csv", 
                     col_select = c(1:9), 
                     col_names = TRUE,
                     col_types = cols(
                       Bee_ID = col_character(),
                       Diapause_NonDiapause = col_character(), 
                       Sex = col_character(), 
                       Date_emerged = col_date("%m/%d/%y"), 
                       Thorax_length_mm = col_double(), 
                       Assay_number = col_character(), 
                       C = col_double(), 
                       P = col_double(), 
                       CP = col_double()
                     ))
problems(oxstress)
```

## Summary stats

```{r sum-1}
hist(oxstress$CP)
```

```{r sum-2}
oxstress %>% 
  group_by(Diapause_NonDiapause) %>% 
  summarise(mean_CP = mean(CP), 
            se_CP = sd(CP)/sqrt(n()), 
            mean_size = mean(Thorax_length_mm), 
            se_size = sd(Thorax_length_mm)/sqrt(n()), 
            n = n()
            )
```

Diapausing females have 26% higher levels of protein carbonylation. 
Diapausing females are also 4% larger than nondiapausing females.

```{r sum-3}
qqp(oxstress$CP, "norm")
ad.test(oxstress$CP)
```


## Analysis

```{r stats-1}
m1 <- lmer(CP ~ Diapause_NonDiapause + Thorax_length_mm + (1|Assay_number), 
           data = oxstress)
summary(m1)
Anova(m1)
AIC(m1)
```

```{r stats-2}
m0 <- lmer(CP ~ 0 + (1|Assay_number), data = oxstress)
summary(m0)
anova(m1, m0)
```

## Plot

```{r plot}
ggplot(oxstress, aes(x = Diapause_NonDiapause, y = CP, 
                     fill = Diapause_NonDiapause)) + 
  geom_boxplot(alpha = 0.8, outlier.color=NA) + 
  geom_jitter(position = position_jitter(width = 0.20)) + 
  geom_signif(data = oxstress, stat = "signif", position = "identity", 
              comparisons = list(c("D", "ND")), map_signif_level = TRUE, 
              annotations = "***", textsize = 8) + 
  coord_cartesian(ylim = c(0.1, 0.7)) +
  theme_classic() + 
  scale_fill_manual(values = c("#94D2BD","#D84727")) + 
  theme(legend.position = "none", 
        axis.title=element_text(size=16), axis.text=element_text(size=14)) +
  xlab("") + 
  scale_x_discrete(labels = c("Diapause", "Non-diapause")) + 
  ylab("Protein carbonyl / Total protein (nmol/mg)")
ggsave("oxstress.png", width = 7, height = 5)
```

There is an outlier in the non-diapause group. This is MRE053. 
It is extremely small, based on thorax width, and Addie had a note that the 
protein precipitate was hardly visible for 53 and 54. Therefore, it is likely 
that the CP value is exaggerated due to the very small amount of protein in 
this individual due to its small size.


## bookkeeping

```{r books}
sessionInfo()
citation("tidyverse")
citation("car")
citation("nortest")
citation("lme4")
citation("ggsignif")
```